Red States = Red Flags?

### 2024-02-13 paper/paper.qmd
update qmd and pdf
changes to pdf and scripts
Added graphs to paper output and changed the path for various scripts

{r}
#| include: false
#| warning: false
#| message: false

@@

@@

Model InformationDependent variable: birth rate per 1000 populationIndependent variables:State min wage Abortion delay Child enforcement expenditure Political affiliation Education

@@

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the rstanarm package of @rstanarm. We use the default priors from rstanarm.

@@

{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

@@

{r}
#INTREST DISTRO OVER TIME GRAPHS

@@

The District of Columbia (DC) exhibits the largest increase, with its minimum wage rising by $7.85, from $6.15 in 2001 to $14.00 in 2019. Arizona (AZ) and Colorado (CO) also show substantial increases of $6.85 each.

Georgia (GA) and Wyoming (WY) stand out as the only states with no change in minimum wage, maintaining a rate of $5.15 across the years. It is worth noting that this rate is equal to the federal minimum wage between 1997 - 2007, suggesting that these states have not adjusted their minimum wage above the federal floor set at that time.

{r}
# MINIMUM WAGE GRAPH

@@

{r}
#relative average # of births in states delay

#relative average # of births in states w/o delay

@@

@@

{r}
# POLITICAL AFFILIATION MAP

@@

{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

@@

{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

>>>>

```{r echo=FALSE, message=FALSE}
#| include: false
#| warning: false
#| message: false

@@

```

@@

Model Information Dependent variable: birth rate per 1000 population
Independent variables:State min wage, Abortion delay, Child enforcement
expenditure, Political affiliation

Education

@@

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the rstanarm package of @rstanarm.
We use the default priors from rstanarm.

@@

```{r echo=FALSE, message=FALSE}
#| echo: false
#| eval: true
#| warning: false
#| message: false

@@

{r} #\| echo: false #\| eval: true #\| label: tbl-modelresults #\|
tbl-cap: "Explanatory models of flight time based on wing width and wing
length" #\| warning: false

@@

```{r echo=FALSE, message=FALSE}
library(haven)
library(tidyverse)
library(tidyr)

# Importing the CSV files
numbirths_2001_2019 <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv")
births_educ_race_age6 <- read_csv("../data/analysis_data/decomp/births_educ_race_age6.csv")
numbirths_educ_2044 <- read_csv("../data/analysis_data/educ/numbirths_educ_2044.csv")
state_births_04_19 <- read_csv("../data/analysis_data/map/state_births_04_19.csv")

# Display the first few rows of each data-frame to verify the imports
view(numbirths_2001_2019)
view(births_educ_race_age6)
view(numbirths_educ_2044)
view(state_births_04_19)

### Number of Births ### 
colnames(numbirths_2001_2019)

average_births_by_state <- numbirths_2001_2019 %>%
  filter(year >= 2001, year <= 2019) %>%
  group_by(stname) %>%
  summarize(AverageBirths = mean(numbirth1544, na.rm = TRUE)) %>%
  ungroup()
view(average_births_by_state)

ggplot(average_births_by_state, aes(y = reorder(stname, AverageBirths), x = AverageBirths)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Average Number of Births by State (2001-2019)",
       y = "State",
       x = "Average Number of Births") +
  theme(axis.text.y = element_text(angle = 0)) 
```

@@

The District of Columbia (DC) exhibits the largest increase, with its
minimum wage rising by \$7.85, from \$6.15 in 2001 to \$14.00 in 2019.
Arizona (AZ) and Colorado (CO) also show substantial increases of \$6.85
each.

Georgia (GA) and Wyoming (WY) stand out as the only states with no
change in minimum wage, maintaining a rate of \$5.15 across the years.
It is worth noting that this rate is equal to the federal minimum wage
between 1997 - 2007, suggesting that these states have not adjusted
their minimum wage above the federal floor set at that time.

```{r echo=FALSE, message=FALSE, results='asis'}
library(knitr)
library(readxl)
library(dplyr)
library(ggplot2)

# Load Data 
welfare_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data")

# Retrieving data for state minimum wage for 2001 - 2019
filtered_minwage <- welfare_data %>%
  select(state_name, year, `State Minimum Wage`) %>%
  filter(year >= 2001, year <= 2019)

##### State Minimum Wage #####
# Filter data for 2001 and 2019
minwage_2001 <- welfare_data %>%
  filter(year == 2001) %>%
  select(state_name, year, `State Minimum Wage`)

minwage_2019 <- welfare_data %>%
  filter(year == 2019) %>%
  select(state_name, year, `State Minimum Wage`)

# Rename the `State Minimum Wage` columns to reflect the year
minwage_2001 <- minwage_2001 %>%
  rename(MinWage2001 = `State Minimum Wage`)

minwage_2019 <- minwage_2019 %>%
  rename(MinWage2019 = `State Minimum Wage`)

# Merge the 2001 and 2019 data by state_name
minwage_change <- left_join(minwage_2001, minwage_2019, by = "state_name")

# Calculate the change in minimum wage from 2001 to 2019
minwage_change <- minwage_change %>%
  mutate(ChangeInMinWage = MinWage2019 - MinWage2001) %>%
  select(state_name, MinWage2001, MinWage2019, ChangeInMinWage)

kable(minwage_change, format = "markdown", caption = "Change in State Minimum Wage from 2001 to 2019")

```

@@

#relative average \# of births in states delay

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(readr)

# Determine which states had a delay or no delay policy 
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  filter(year >= 2001, year <= 2019) %>%
  select(delay, stname, year)

# Load the population data from the Excel file
population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
  select(state_name, year, Population) %>%
  filter(year >= 2001, year <= 2019)

numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
  rename(births = numbirth1544) %>%
  filter(year >= 2001, year <= 2019)

# Ensure both datasets have the 'stname' column for joining
population_data <- population_data %>%
  rename(stname = state_name)

# Merge the number of births data with the population data
births_population_data <- numbirths_data %>%
  left_join(population_data, by = c("year", "stname"))

# Calculate the relative average number of births per 1000 people
births_population_data <- births_population_data %>%
  mutate(relative_births = (births / Population) * 1000)

# Load the abortion policy data
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  select(year, stname, delay)

# Merge the relative births data with the abortion policy data
combined_data <- abortion_policy %>%
  left_join(births_population_data, by = c("year", "stname"))

# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")

# Filter the combined_data for states with no delay
combined_data_no_delay <- combined_data %>%
  filter(stname %in% states_no_delay)

# Filter the combined_data for states with delay
combined_data_with_delay <- combined_data %>%
  filter(stname %in% states_with_delay)

# Plot for states with policy delay
ggplot(data = combined_data_with_delay) +
  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
  labs(title = "Relative Average Number of Births in States With Delay (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People",
       color = "State") +
  theme_minimal()

```

#relative average \# of births in states w/o delay

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(readr)

# Determine which states had a delay or no delay policy 
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  filter(year >= 2001, year <= 2019) %>%
  select(delay, stname, year)

# Load the population data from the Excel file
population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
  select(state_name, year, Population) %>%
  filter(year >= 2001, year <= 2019)

numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
  rename(births = numbirth1544) %>%
  filter(year >= 2001, year <= 2019)

# Ensure both datasets have the 'stname' column for joining
population_data <- population_data %>%
  rename(stname = state_name)

# Merge the number of births data with the population data
births_population_data <- numbirths_data %>%
  left_join(population_data, by = c("year", "stname"))

# Calculate the relative average number of births per 1000 people
births_population_data <- births_population_data %>%
  mutate(relative_births = (births / Population) * 1000)

# Load the abortion policy data
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  select(year, stname, delay)

# Merge the relative births data with the abortion policy data
combined_data <- abortion_policy %>%
  left_join(births_population_data, by = c("year", "stname"))

# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")

# Filter the combined_data for states with no delay
combined_data_no_delay <- combined_data %>%
  filter(stname %in% states_no_delay)

# Filter the combined_data for states with delay
combined_data_with_delay <- combined_data %>%
  filter(stname %in% states_with_delay)

# Plot for states with no policy delay
ggplot(data = combined_data_no_delay) +
  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
  labs(title = "Relative Average Number of Births in States Without Delay (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People",
       color = "State") +
  theme_minimal()

```

#relative birth rates in transition states

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(readr)

# Determine which states had a delay or no delay policy 
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  filter(year >= 2001, year <= 2019) %>%
  select(delay, stname, year)

# Load the population data from the Excel file
population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
  select(state_name, year, Population) %>%
  filter(year >= 2001, year <= 2019)

numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
  rename(births = numbirth1544) %>%
  filter(year >= 2001, year <= 2019)

# Ensure both datasets have the 'stname' column for joining
population_data <- population_data %>%
  rename(stname = state_name)

# Merge the number of births data with the population data
births_population_data <- numbirths_data %>%
  left_join(population_data, by = c("year", "stname"))

# Calculate the relative average number of births per 1000 people
births_population_data <- births_population_data %>%
  mutate(relative_births = (births / Population) * 1000)

# Load the abortion policy data
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  select(year, stname, delay)

# Merge the relative births data with the abortion policy data
combined_data <- abortion_policy %>%
  left_join(births_population_data, by = c("year", "stname"))

# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")

# Filter the combined_data for states with no delay
combined_data_no_delay <- combined_data %>%
  filter(stname %in% states_no_delay)

# Filter the combined_data for states with delay
combined_data_with_delay <- combined_data %>%
  filter(stname %in% states_with_delay)

# States and years of policy implementation
policy_changes <- data.frame(
  stname = c("AZ", "GA", "MN", "NC", "OK", "TX", "WV"),
  year_implemented = c(2009, 2005, 2003, 2011, 2005, 2003, 2003)
)

# Filter combined_data
transition_states_data <- combined_data %>%
  filter(stname %in% policy_changes$stname)

# Creating the graph
ggplot(data = transition_states_data) +
  geom_line(aes(x = year, y = relative_births, group = stname, color = stname)) +
  geom_vline(data = policy_changes, aes(xintercept = year_implemented, color = stname), linetype="dashed") +
  geom_text(data = policy_changes, aes(x = year_implemented, y = Inf, label = year_implemented), vjust = -0.5, color = "black") +
  scale_x_continuous(breaks = seq(2001, 2019, by = 2)) +
  labs(title = "Transition to Policy Delay and Relative Average Number of Births (2001-2019)",
       x = "Year",
       y = "Relative Average Number of Births per 1000 People",
       color = "State") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

@@

```{r echo=FALSE, message=FALSE}
library(readxl)
library(dplyr)
library(readr)

# Determine which states had a delay or no delay policy 
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  filter(year >= 2001, year <= 2019) %>%
  select(delay, stname, year)

# Load the population data from the Excel file
population_data <- read_excel("../data/UKCPR_National_Welfare_Data_Update_020623.xlsx", sheet = "Data") %>%
  select(state_name, year, Population) %>%
  filter(year >= 2001, year <= 2019)

numbirths_data <- read_csv("../data/analysis_data/annual_policy/numbirths_2001_2019.csv") %>%
  rename(births = numbirth1544) %>%
  filter(year >= 2001, year <= 2019)

# Ensure both datasets have the 'stname' column for joining
population_data <- population_data %>%
  rename(stname = state_name)

# Merge the number of births data with the population data
births_population_data <- numbirths_data %>%
  left_join(population_data, by = c("year", "stname"))

# Calculate the relative average number of births per 1000 people
births_population_data <- births_population_data %>%
  mutate(relative_births = (births / Population) * 1000)

# Load the abortion policy data
abortion_policy <- read_dta("../data/analysis_data/annual_policy/policyvars01_19.dta") %>%
  select(year, stname, delay)

# Merge the relative births data with the abortion policy data
combined_data <- abortion_policy %>%
  left_join(births_population_data, by = c("year", "stname"))

# Define the groups based on predefined 'states_no_delay' and 'states_with_delay'
states_no_delay <- c("AK", "AL", "CA", "CO", "CT", "DC", "DE", "FL", "HI", "IA", "IL", "MA", "MD", "ME", "MT", "NH", "NJ", "NM", "NV", "NY", "OR", "RI", "TN", "VT", "WA", "WY")
states_with_delay <- c("AR", "ID", "IN", "KS", "KY", "LA", "MI", "MO", "MS", "ND", "NE", "OH", "PA", "SC", "SD", "UT", "VA", "WI")

# Filter the combined_data for states with no delay
combined_data_no_delay <- combined_data %>%
  filter(stname %in% states_no_delay)

# Filter the combined_data for states with delay
combined_data_with_delay <- combined_data %>%
  filter(stname %in% states_with_delay)

# Step 1: Select states for comparison
states_comparison <- combined_data %>%
  filter(stname %in% c("CA", "TX", "GA", "NC", "FL"))

# Step 2: Calculate descriptive statistics and visualize trends
# For California (no policy change) and Texas (policy change in 2003)
ca_tx_data <- states_comparison %>%
  filter(stname %in% c("CA", "TX"))

# Calculate summary statistics for CA and TX before and after 2003
ca_tx_summary <- ca_tx_data %>%
  group_by(stname) %>%
  summarise(
    mean_before = mean(relative_births[year < 2003], na.rm = TRUE),
    mean_after = mean(relative_births[year >= 2003], na.rm = TRUE)
  )

# For Georgia (policy change in 2005), North Carolina (policy change in 2011), and Florida (no policy change)
ga_nc_fl_data <- states_comparison %>%
  filter(stname %in% c("GA", "NC", "FL"))

# Calculate summary statistics for GA and NC before and after policy changes, and compare with FL
ga_nc_fl_summary <- ga_nc_fl_data %>%
  group_by(stname) %>%
  summarise(
    mean_before = mean(relative_births[year < if_else(stname == "GA", 2005, if_else(stname == "NC", 2011, year))], na.rm = TRUE),
    mean_after = mean(relative_births[year >= if_else(stname == "GA", 2005, if_else(stname == "NC", 2011, year))], na.rm = TRUE)
  )

# Step 3: Visualize trends
ggplot(states_comparison, aes(x = year, y = relative_births, color = stname)) +
  geom_line() +
  geom_vline(xintercept = c(2003, 2005, 2011), linetype = "dashed", color = "black") +
  labs(title = "Relative Birth Rates in Selected States (2001-2019)", y = "Relative Births per 1000 People", x = "Year") +
  theme_minimal()

```

@@

{r} \# POLITICAL AFFILIATION MAP

@@

{r} #\| eval: true #\| echo: false #\| message: false #\| warning: false
#\| label: fig-ppcheckandposteriorvsprior #\| layout-ncol: 2 #\|
fig-cap: "Examining how the model fits, and is affected by, the data"
#\| fig-subcap: \["Posterior prediction check", "Comparing the posterior
with the prior"\]

@@

{r} #\| echo: false #\| eval: true #\| message: false #\| warning: false
#\| label: fig-stanareyouokay #\| fig-cap: "Checking the convergence of
the MCMC algorithm" #\| fig-subcap: \["Trace plot", "Rhat plot"\] #\|
layout-ncol: 2
